Supplement to: Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data

### 2022-10-11 analysis/paper/paper.Rmd
reading submission

<!-- This can hinder the extrapolation from individual sampling sites toward predicting effect across entire urban areas and tree stocks, especially where studies rely on labor-intensive methods which are limited logistically by sampling effort, reducing sample sizes, and species as well as spatial coverage. -->
<!-- This is exacerbated by a lack of co-located environmental variables (i.e. measured in situ) at pertinent spatial scales, for instance, as noted by @wohlfahrt2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity when applying space-for-time substitutions. -->
<!-- Space-for-time substitutions and time series comparisons between and within locations are a common approach (cf. studies above) to generate inferences in observational (rather than treatment-control) studies, where manipulations are costly or logistically unfeasible due to time and/or financial constraints. -->
<!-- However, they require  accounting for confounding factors specific to trees' environments, such as street characteristics, development intensity, available soil volume, etc. -->
<!-- While several of the aforementioned studies applied these approaches to quantify temperature and excess heat on growth, they typically compare trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatio-temporal variability in location-specific factors noted above. -->
Several of the aforementioned studies applied space-for-time substitutions or time series comparisons to quantify temperature and excess heat on growth for a given region.
However, they typically compared trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatial and/or temporal variability in location-specific factors noted above.
This can hinder the extrapolation from individual sampling sites toward predicting effects across entire urban areas and tree stocks, especially when studies rely on labor-intensive methods, 
<!-- which are limited logistically by sampling effort,  -->
reducing sample sizes and coverage of species and space.
A lack of co-located environmental variables (i.e. measured *in situ*) at pertinent spatial scales can further exacerbate these limitations, for instance, as noted by @wohlfahrt2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity.
It is thus likely that the varying and even contrasting growth responses observed for urban trees across and within studies are at least modulated by some confounding factors, making the attribution to a single driver, such as excess heat, more difficult and possibly less accurate.

@@

This approach is considered conservative (see supplementary information), yet all analyses were carried forward with the unfiltered and filtered data - no considerable differences were found, thus subsequent sections are based on the filtered data.
<!-- which were processed with the bespoke software datacleanr by Hurley et al., submitted), -->
<!-- Characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via https://daten.berlin.de and curated by the city's government. -->
<!-- The data used from the inventory includes species, location, age, and diameter at breast height (dbh). -->
Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.
Final samples applied in models were smaller, following the availability of ancillary data for a given observation, and limited to a maximum age of 125 years to increase confidence in reported values, and ultimately model estimates.

@@

## Conclusion

@@

AGH and IH were supported through the Helmholtz-Climate-Initiative (HI-CAM), funded by the Helmholtz Initiative and Networking Fund; the authors are responsible for the content of this publication.

# Author Contributions

>>>>

<!-- This can hinder the extrapolation from individual sampling sites toward predicting effect across entire urban areas and tree stocks, especially where studies rely on labor-intensive methods which are limited logistically by sampling effort, reducing sample sizes, and species as well as spatial coverage. -->
<!-- This is exacerbated by a lack of co-located environmental variables (i.e. measured in situ) at pertinent spatial scales, for instance, as noted by @wohlfahrt2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity when applying space-for-time substitutions. -->
<!-- Space-for-time substitutions and time series comparisons between and within locations are a common approach (cf. studies above) to generate inferences in observational (rather than treatment-control) studies, where manipulations are costly or logistically unfeasible due to time and/or financial constraints. -->
<!-- However, they require  accounting for confounding factors specific to trees' environments, such as street characteristics, development intensity, available soil volume, etc. -->
<!-- While several of the aforementioned studies applied these approaches to quantify temperature and excess heat on growth, they typically compare trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatio-temporal variability in location-specific factors noted above. -->
Multiple of the aforementioned studies applied space-for-time substitutions or time series comparisons to quantify temperature and excess heat on growth for a given region.
However, they typically compared trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatial and/or temporal variability in location-specific factors noted above.
This can hinder the extrapolation from individual sampling sites toward predicting effects across entire urban areas and tree stocks, especially when studies rely on labor-intensive methods, 
<!-- which are limited logistically by sampling effort,  -->
reducing sample sizes and coverage of species and space.
A lack of co-located environmental variables (i.e. measured *in situ*) at pertinent spatial scales can further exacerbate these limitations, for instance, as noted by @wohlfahrt2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity.
It is thus likely that the varying and even contrasting growth responses observed for urban trees across and within studies are at least modulated by some confounding factors, making the attribution to a single driver, such as excess heat, more difficult and possibly less accurate.

@@

Similar cut-off approaches are frequently applied in allometric and trait databases [e.g., @kattge2020]. We considered this approach (i.e., $MAD \cdot 7$) conservative, and necessary given the presence of clearly erroneous data.
The approach and impact thereof are visualized in the supplementary information (Section Data quality control).
<!-- which were processed with the bespoke software datacleanr by Hurley et al., submitted), -->
<!-- Characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via https://daten.berlin.de and curated by the city's government. -->
<!-- The data used from the inventory includes species, location, age, and diameter at breast height (dbh). -->
Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.
Final samples applied in models were smaller, following the availability of ancillary data for a given observation, and limited to a maximum age of 125 years to increase confidence in reported values, and ultimately model estimates.

@@

# Conclusion

@@

AGH and IH were supported through the Helmholtz-Climate-Initiative (HI-CAM), funded by the Helmholtz Initiative and Networking Fund; the authors are responsible for the content of this publication.
We thank Prof Dr Christoph Schneider for providing dendrochronological data and pertinent discussion on its use.

### 2022-10-11 analysis/paper/supplemental_information.Rmd
adjusting eqs
reading submission

---
title: "Supplement to: Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data"
author:
  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$, ...
  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Germany*
  - "* Corresponding author: hurley@gfz-potsdam.de"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::html_document2:
      fig_caption: yes
      toc: true
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      toc: yes
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}
bibliography: 
  - uhi_references.bib
  - 04_gfz_UHI.bib
  - 07_phenoTram.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  **This document outlines the rationale for an analysis of tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin using an extensive, publicly available data set. It introduces preliminary results and provides an outlook for up-coming and potential work.**
keywords: |
  Uferbaeume; Anlagenbaeume; Strassenbaeume; 
highlights: |
  We will see. 
---

@@

# Overview

- Upload full code with DOI
- datacleanr imgs
- diagnostic lm imgs
- spatial comparison (Moran's I)

<!-- # Acknowledgements -->

<!-- <!-- The following line inserts a page break  --> 
<!-- `r pgbreak()` -->

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

`r pgbreak()`

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

>>>>

---
title: "Supplement to: Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data"
author:
  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$
  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Germany*
  - "* Corresponding author: hurley@gfz-potsdam.de"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::html_document2:
      fig_caption: yes
      toc: true
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}
bibliography: 
  - uhi_references.bib
  - 04_gfz_UHI.bib
  - 07_phenoTram.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style

---

@@

# Code and reproducibility

The code to reproduce the paper and its analyses, as well as this supplement, are found at [https://github.com/the-Hull/berlin.trees/](), with permanently archived versions at **DOI**.
The project relies on the workflow management tool `drake`, where all analyses steps are maintained within the file `berlin.trees/R/plan.R`, with necessary functions in `berlin.trees/R/functions.R`.
The code downloads most necessary files to execute the analyses.
However, some files were too large for tracking with version control (i.e., `git` on GitHub); thus, should any issues arise with executing the code, please contact the corresponding author.

Further information on `drake` is available at:  
- William Michael Landau, (2018). The drake R package: a pipeline toolkit for
  reproducibility and high-performance computing. **Journal of Open Source Software, 3(21)**,
  550, https://doi.org/10.21105/joss.00550

# Data quality control for street tree inventory

Manual data processing for quality control was done with a bespoke software  `datacleanr` by [@hurley2022], where obvious outliers or clearly interpolated data were removed.
The latter was deemed necessary, as several observations in multiple city districts were derived by linear relationships (i.e., straight-line), which do not capture the ontogenic growth dynamics of trees, and leave no variation related to variables other than age.
All of these operations were recorded, and can be viewed in the repository [https://github.com/the-Hull/berlin.trees/]() and reproduced via the supplementary code:

- exploratory and diagnostic plots for age and diameter relationships: `berlin.trees/analysis/figures/eda/`
- `datacleanr` cleaning operations: `berlin.trees/analysis/data/raw_data/tree_splits`

```{r fig-diag-dcr, fig.cap = "Manual data inspection and cleaning with the R package `datacleanr`. This example for the genus *Tilia* in the district Spandau clearly shows linearly interpolated age-diameter observations. These were manually removed with utmost diligence to retain non-interpolated data. Note, that the supplementary code includes files to view and reproduce the results of manual cleaning in *analysis/data/raw_data/tree_splits/*.", fig.width = 14, fig.height = 14, out.width="100%"}

knitr::include_graphics(here::here("analysis/figures/diagnostic_00_datacleanr-Tilia-district_spandau.png"))
```

Lastly, observations with unlikely diameter-age combinations were identified via the residuals of a generalized linear model between diameter and age with a Gamma log-link distribution: if individual residuals exceeded seven times the median absolute deviation of all residuals, they were removed.
The median absolute deviation (MAD) is comparable to the inter-quartile range, yet more robust to outliers:

\begin{equation}
  MAD = median(\|X_i - median(X)\|) 
  (\#eq:mad)
\end{equation}

```{r fig-diag-lm, fig.cap = "Observations with unlikely diameter-age combinations ($dbh_{cm}~vs.~STANDALTER$) were identified via the residuals of a generalized linear model between diameter and age with a Gamma log-link distribution for every species included in the study. Where individual residuals exceeded seven times the median absolute deviation of all residuals, they were removed. The line is the fitted model, and colors represent retained or removed observations (see legend). See main text (Section Street Trees) for further explanation.", fig.width = 14, fig.height = 14, out.width="100%"}

knitr::include_graphics(here::here("analysis/figures/diagnostic_01_model_data.png"))
```

# Spatial auto-correlation

Exploratory analyses of (initial) model resiudals showed spatial auto-correlation was present in the structure of the inventory data (identified by correlation of x and y coordinates with residuals).
Note that this was expected given that trees planted on the same road (and/or neighborhood, district) may share several features, such as planting time, species, adjacent urban fabric. 
Thus, we included model features (i.e., dependent variables) in the model formulation to account for spatial auto-correlation. 
This was achieved with isometric smoothing splines with basis derived from both x and y co-ordinates (i.e., interaction terms in the generalized additive model). 
We used Morans' I to calcualte spatial dependency with the function `Moran.I()` from the `R` package `ape`.

$$
I = \frac{n}{S_0} \cdot \frac{\sum_{i=1}^{n} \sum_{j=1}^{n} w_{ij}(y_{i} - \bar{y})(y_{j} - \bar{y})} { \sum_{i=1}^{n} (y_{i} - \bar{y})^2}
$$
where $y_{i}$ are individual observations, $w_{ij}$ are distance weights, $n$ are the totall number of observations, and $S_{0} = \sum_{i=1}^{n} \cdot \sum_{j=1}^{n} w_{ij}$.

Details on `ape` can be found in:

- *Paradis E. & Schliep K. 2019. ape 5.0: an environment for modern phylogenetics and evolutionary analyses in R. **Bioinformatics 35**: 526-528.*

Due to computational limitations, we split the data set into several groups using a spatial grid from the Berlin authorities, based on $2 \cdot 2~km$ cells, found at:  
- [https://fbinter.stadt-berlin.de/fb/atom//DOP/Blattschnitt2x2km.gif](),

with further details at:  
- [https://fbinter.stadt-berlin.de/fb/berlin/service_intern.jsp?id=a_luftbild2010_cir@senstadt&type=FEED]() (accessed at 2022-09-29),

with the resulting splits below:

```{r fig-grid, fig.cap = "Grid for splitting observations into discrete groups. Regions are based on $2 \\cdot 2~km$ grid cells and grouped based on Berlin's data acquisition scheme for aerial images", echo = FALSE, out.width='50%'}

grid <- sf::st_read(here::here("analysis/data/raw_data/spatial_ancillary/grid_2x2.geojson"), quiet = TRUE)

library(ggplot2)
ggplot() +
  geom_sf(data = grid, aes(fill = grid_region, color = grid_region)) +
  scale_fill_brewer(palette = 'Set3') +
  scale_color_brewer(palette = 'Set3') +
  theme_minimal() +
  labs(title = "Gridded Berlin regions (2x2 km)", color = "Grid Region", fill = "Grid Region")

```

We observed a considerable reduction of Moran's I (i.e., spatial auto-correlation) by including spatial terms in the model structure (i.e., x and y coordinates). 
Given the large sample sizes available for this study we considered this approach (i.e., spatial terms) satisfactory to ensure the inference of model coefficients was not violated due to dependence of residuals.

```{r fig-moran, echo = FALSE, fig.cap = "Comparison of Moran's I with and without spatial terms included in the model structure as dependent variable (bar color) between two model structures (additive on left, and interaction structures for temperature on the right).", out.width='100%'}

knitr::include_graphics(here::here("analysis/figures/fig-gam-spatial_moran_comparison_blocks.png"))

```
